### 2021-04-15 response.Rmd

---
#title: "Quantifying the Impact of Nasopharyngeal Specimen Quality of SARS-CoV-2 Test Performance"
#author: "Brendan J. Kelly, MD, MS"
bibliography: bib/references.bib
csl: csl/ofid.csl
output: html_document
css: style_helvetica_double.css
---

```{r setup, echo=FALSE, eval=TRUE, include=FALSE, autodep=TRUE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(tidybayes)

```

`r format(Sys.Date(), "%d %B %Y")`

<br>

To the editor:

Thank you for your review of our manuscript, titled "Quantifying the Impact of Nasopharyngeal Specimen Quality on SARS-CoV-2 Test Performance." We have made substantial revisions to the manuscript and added new analyses in response to the reviewers' comments. We believe the manuscript is significantly improved as a result of these changes. We have attached the new manuscript with changes highlighted, and below we have summarised our responses to the reviewers' comments, including the locations of edits:

<br>

```{r response_table, echo=FALSE, eval=TRUE, include=TRUE, message=FALSE, warning=FALSE, error=FALSE, autodep=TRUE}

library(tidyverse)
library(gt)

revision_tab <- read_tsv("./revision/ofid_review_response.tsv")

revision_tab %>%
  gt::gt() %>%
  gt::opt_table_lines(extent = "all")

```

<br>

We believe that the attached manuscript will be of interest to the infectious diseases, clinical microbiology, and epidemiology communities, particularly in light of growing interest in using SARS-CoV-2 Ct values to guide infection control and public health interventions. All authors have reviewed and approved the attached manuscript, and we report no conflicts of interest. The manuscript is not under consideration at any other journal. Please contact me (brendank@pennmedicine.upenn.edu) in case of any questions.

Many thanks for your consideration,

<br>

Brendan J. Kelly, MD, MS
Assistant Professor of Medicine & Epidemiology  
Division of Infectious Diseases  
Perelman School of Medicine  
University of Pennsylvania

### 2021-04-16 manuscript.Rmd
Fix typos.

* __Statistical Methods__: Data were organized using R statistical software version 3.6.1 [@R_Core_Team2018-xo], and plots generated using the “ggplot2” package [@Wickham2016-eh]. Where β-actin and SARS-CoV-2 were not detected, Ct values were imputed as 40 cycles. We examined (1) the linear relationship between β-actin and SARS-CoV-2 Ct values, as well as (2) the impact of β-actin Ct on SARS-CoV-2 detection using Bayesian linear and generalized-linear mixed effects models, which were fit using Stan Hamiltonian Monte Carlo (HMC) version 2.21, via the "brms" package with default weakly-informative priors [@Carpenter2017-ec; @Burkner2017-yr]. Mixed effects models were fit in the same manner, with random slopes and intercepts according to disease severity (assessed by radiography or oxygen saturation) to evaluate the impact of disease severity on the relationship. Prior predictive modeling was performed, and models were fit with 4 chains of 1000 iterations, confirmed with HMC diagnostics (no divergent iterations, Rhat statistic < 1.1 for all parameters, and E-BFMI > 0.2) [@McElreath2016-sf; @Gabry2019-le; @Gelman2020-yu]. We examined parameter distributions at 50%, 80%, and 95% posterior credible intervals to understand the relationship between exposure and outcome variables. The posterior credible intervals estimate the probability that the true parameter value lies within the interval, based on the observed data.

@@

* __SARS-CoV-2 Detection and Cycle Threshold Range__: Of 1282 tested specimens, 134 were found to have detectable SARS-CoV-2 within 40 cycles of PCR. <span style='background-color:lightcyan'>Among these specimens, median (interquartile range (IQR)) SARS-CoV-2 Ct was 27.9 (20.4 to 32.9). No secular trend was observed between calendar time from local onset of COVID-19 cases and SARS-CoV-2 Ct values during the study period (Spearman correlation 0.18). **Table 1** summarises the SARS-CoV-2 Ct values, as well as those of the specimen quality control, β-actin, and RT-PCR reaction control, MS2 RNA. **Supplemental Figure 1** depicts the distribution of each target and the relationships between their distributions.</span>

@@

* __Impact of Disease Severity on Relationship Between β-Actin and SARS-CoV-2 Cycle Threshold__: Considering the possibility that the observed association between SARS-CoV-2 and β-actin Ct values is confounded by respiratory illness severity, we evaluated the relationship between β-actin Ct and independent markers of respiratory illness. Oxygen saturation data were available for 425 (33.1%) subjects; chest CT imaging was available for 108 (8.4%) subjects. <span style='background-color:lightcyan'>At least one measure of disease severity was available for 38 (28.3%) of the cases of detectable SARS-CoV-2.</span> Linear regression relating β-actin Ct values to oxygen saturation revealed no significant association with linear regression slope -0.04 (95%CI -0.273 to 0.182). Similarly, we found that the presence of parenchymal lung disease on chest CT radiography reports had no significant association with β-actin Ct values, with a linear regression slope 0.428 (95%CI -0.655 to 1.61). Mixed effects models revealed that the observed association between β-actin Ct values and probability of SARS-CoV-2 detection persists across all levels of respiratory illness, whether assessed by oxygen saturation or chest radiography (**Figure 2**). These analyses of independent markers of severe respiratory disease suggest that it is NP specimen quality, not disease severity, that drives the association between SARS-CoV-2 and β-actin Ct values. <span style='background-color:lightcyan'>To further corroberate these findings, we repeated the analysis, comparing specimens collected from inpatient (n = 858) versus outpatient (n = 424) locations. We found the same association between β-actin Ct values and probability of SARS-CoV-2 detection across both specimen collection sites.</span>

>>>>

* __Statistical Methods__: Data were organized using R statistical software version 3.6.1 [@R_Core_Team2018-xo], and plots generated using the “ggplot2” package [@Wickham2016-eh]. Where β-actin and SARS-CoV-2 were not detected, Ct values were imputed as 40 cycles. We examined (1) the linear relationship between β-actin and SARS-CoV-2 Ct values, as well as (2) the impact of β-actin Ct on SARS-CoV-2 detection using Bayesian linear and generalized-linear mixed effects models, which were fit using Stan Hamiltonian Monte Carlo (HMC) version 2.21, via the "brms" package with default weakly-informative priors [@Carpenter2017-ec; @Burkner2017-yr]. Mixed effects models were fit in the same manner, with random slopes and intercepts according to disease severity (assessed by radiography or oxygen saturation) to evaluate the impact of disease severity on the relationship. Prior predictive modeling was performed, and models were fit with 4 chains of 1000 iterations, confirmed with HMC diagnostics (no divergent iterations, Rhat statistic < 1.1 for all parameters, and E-BFMI > 0.2) [@McElreath2016-sf; @Gabry2019-le; @Gelman2020-sd]. We examined parameter distributions at 50%, 80%, and 95% posterior credible intervals to understand the relationship between exposure and outcome variables. The posterior credible intervals estimate the probability that the true parameter value lies within the interval, based on the observed data.

@@

* __SARS-CoV-2 Detection and Cycle Threshold Range__: Of 1282 tested specimens, 134 were found to have detectable SARS-CoV-2 within 40 cycles of PCR. <span style='background-color:lightcyan'>Among these specimens, median (interquartile range (IQR)) SARS-CoV-2 Ct was 27.9 (20.4 to 32.9). No secular trend was observed between calendar time from local onset of COVID-19 cases and SARS-CoV-2 Ct values during the study period (Spearman correlation 0.18). **Table 1** summarizes the SARS-CoV-2 Ct values, as well as those of the specimen quality control, β-actin, and RT-PCR reaction control, MS2 RNA. **Supplemental Figure 1** depicts the distribution of each target and the relationships between their distributions.</span>

@@

* __Impact of Disease Severity on Relationship Between β-Actin and SARS-CoV-2 Cycle Threshold__: Considering the possibility that the observed association between SARS-CoV-2 and β-actin Ct values is confounded by respiratory illness severity, we evaluated the relationship between β-actin Ct and independent markers of respiratory illness. Oxygen saturation data were available for 425 (33.1%) subjects; chest CT imaging was available for 108 (8.4%) subjects. <span style='background-color:lightcyan'>At least one measure of disease severity was available for 38 (28.3%) of the cases of detectable SARS-CoV-2.</span> Linear regression relating β-actin Ct values to oxygen saturation revealed no significant association with linear regression slope -0.04 (95%CI -0.273 to 0.182). Similarly, we found that the presence of parenchymal lung disease on chest CT radiography reports had no significant association with β-actin Ct values, with a linear regression slope 0.428 (95%CI -0.655 to 1.61). Mixed effects models revealed that the observed association between β-actin Ct values and probability of SARS-CoV-2 detection persists across all levels of respiratory illness, whether assessed by oxygen saturation or chest radiography (**Figure 2**). These analyses of independent markers of severe respiratory disease suggest that it is NP specimen quality, not disease severity, that drives the association between SARS-CoV-2 and β-actin Ct values. <span style='background-color:lightcyan'>To further corroborate these findings, we repeated the analysis, comparing specimens collected from inpatient (n = 858) versus outpatient (n = 424) locations. We found the same association between β-actin Ct values and probability of SARS-CoV-2 detection across both specimen collection sites.</span>
